\textbf{Differences in COVID-19 vaccination in the province of Ontario across Health Regions and socio-economic strata}

### 2023-06-27 manuscript_v6/05-conclusion.qmd
v6 after SW's comments

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
]{article}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[top=20mm,left=20mm,right=20mm,bottom=20mm]{geometry}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\usepackage{caption}
%line numbers
%\usepackage{mathpazo}
%\usepackage{rotating}
\usepackage{lineno}
\linenumbers

\usepackage[noblocks]{authblk}
\renewcommand*{\Authsep}{, }
\renewcommand*{\Authand}{, }
\renewcommand*{\Authands}{, }
\renewcommand\Affilfont{\small}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Differences in COVID-19 vaccination in the province of Ontario across Health Regions and socio-economic strata},
  pdfauthor={Ariel Mundo Ortiz; Bouchra Nasri},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{\textbf{Differences in COVID-19 vaccination in the province of
Ontario across Health Regions and socio-economic strata}}

\author[1,2,3]{Ariel Mundo Ortiz}
\author[1,2,3,*]{Bouchra Nasri}

\affil[1]{Centre de Recherches Mathématiques, University of Montreal,
Montréal, Canada}
\affil[2]{Department of Social and Preventive Medicine, École de Santé
Publique, University of Montreal, Montréal, Canada}
\affil[3]{Centre de recherche en santé publique, University of Montreal,
Montréal, Canada}
\affil[*]{Corresponding author, \url{bouchra.nasri@umontreal.ca}}

\date{}
\begin{document}
\maketitle
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, boxrule=0pt, interior hidden, frame hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, breakable]}{\end{tcolorbox}}\fi

\hypertarget{abstract}{%
\section{Abstract}\label{abstract}}

The COVID-19 pandemic continues to be a worldwide public health concern.
Although vaccines against this disease were rapidly developed,
vaccination uptake has not been equal across all the segments of the
population. In particular, it has been shown that there have been
differences in vaccine uptake across different segments of the
population. However, there are also differences in vaccination across
geographical areas, which might be important to consider in the
development of future public health vaccination policies. In this study,
we examined the relationship between vaccination status (having received
the first dose of a COVID-19 vaccine), and different socio-economic and
geographical factors. Our results show that between October of 2021 and
January of 2022, individuals from underrepresented communities were
three times less likely to be vaccinated than White/Caucasian
individuals across the province of Ontario in Canada, and that in some
cases, within these groups, individuals in low income brackets had
significantly higher odds of vaccination when compared to their peers in
high income brackets. Finally, we identified significantly lower odds of
vaccination in the Central, East and West Health Regions of Ontario
within certain underrepresented groups. This study shows that there is
an ongoing need to better understand and address differences in
vaccination uptake across diverse segments of the population of Ontario
that the pandemic has largely impacted.

\hypertarget{keywords}{%
\section*{Keywords}\label{keywords}}
\addcontentsline{toc}{section}{Keywords}

Covid-19, vaccination, survey, socio-economic factors, visible
minorities.

\hypertarget{background}{%
\section{Background}\label{background}}

As of May of 2023, there have been 765 million confirmed cases of
COVID-19 around the world, including 6.8 million
deaths\textsuperscript{\protect\hyperlink{ref-WHO-Covid}{1}}. Although
this disease is no longer categorized as a global health emergency by
the World Health Organization
(WHO)\textsuperscript{\protect\hyperlink{ref-rigby2023}{2}}, there is
ongoing concern due to continued transmission, surges in cases and
deaths due to new
variants\textsuperscript{\protect\hyperlink{ref-un2023}{3}}, and
weaknesses in health systems around the world that could be exploited by
a novel virus or another public health emergency in the
future\textsuperscript{\protect\hyperlink{ref-mackey2021}{4}}.

In particular, a major weakness that has received attention during the
pandemic has been related to inequalities in vaccine uptake. The rapid
development of vaccines against COVID-19 initially brought the hope of a
rapid end to the pandemic due to the start of vaccination campaigns in
certain parts of the world toward the end of
2020\textsuperscript{\protect\hyperlink{ref-thelancet2021}{5}--\protect\hyperlink{ref-tanne2020}{8}})
but inequalities in vaccine uptake made these pharmaceutical
interventions ultimately unable to replicate the success of the smallpox
vaccination program, which was crucial to control this
disease\textsuperscript{\protect\hyperlink{ref-kayser2021}{9}}.

This problematic is a multifaceted issue resulting from a combination of
factors, among which are failed public health
measures\textsuperscript{\protect\hyperlink{ref-li2021}{10}}, inequality
in vaccine access between high- and low-income
countries\textsuperscript{\protect\hyperlink{ref-gerretsen2021}{11},\protect\hyperlink{ref-tamey2022}{12}},
and vaccine
hesitancy\textsuperscript{\protect\hyperlink{ref-nafilyan2021}{13}}.
Furthermore, it is well established that this issue has affected in
particular individuals in certain underrepresented groups (e.g., Black,
Asian, or Indigenous) as well as individuals with socio-economic
disadvantages\textsuperscript{\protect\hyperlink{ref-willis2021}{14}--\protect\hyperlink{ref-hussain2022}{20}}.

Reasons given for this inequality have included medical mistrust due to
systemic medical
racism\textsuperscript{\protect\hyperlink{ref-stoler2021}{16},\protect\hyperlink{ref-mosby2021}{21}},
mistrust in
vaccines\textsuperscript{\protect\hyperlink{ref-willis2021}{14}}, and
the influence of conspiracy
theories\textsuperscript{\protect\hyperlink{ref-mosby2021}{21}--\protect\hyperlink{ref-freeman2020}{23}}.
However, it is important also to consider that vaccination uptake can be
influenced by geographical (spatial) factors. In this regard,
differences in COVID-19 vaccination rates have been associated with
varied regional attitudes towards
vaccination\textsuperscript{\protect\hyperlink{ref-malik2020}{24}},
spatial differences in vaccine access and supply, vaccination location
availability, and lack of prioritization of areas where vulnerable
groups
reside\textsuperscript{\protect\hyperlink{ref-bogoch2022}{7},\protect\hyperlink{ref-nguyen2021}{25}}.
Other studies have also shown heterogeneity in vaccine uptake within
small governmental administrative units such as
counties\textsuperscript{\protect\hyperlink{ref-mollalo2021}{26}--\protect\hyperlink{ref-bhuiyan2022}{29}},
and that accounting for geographical differences in vaccination can help
predict patterns of booster
uptake\textsuperscript{\protect\hyperlink{ref-wood2022}{30}}.

However, such analyses have been carried out mostly in territories
outside of Canada, where available studies have been focused in certain
cities (such as
Toronto\textsuperscript{\protect\hyperlink{ref-choi2021}{31}}, or
Montreal\textsuperscript{\protect\hyperlink{ref-mckinnon2021}{32}}), or
have explored differences at a province-wide
level\textsuperscript{\protect\hyperlink{ref-guay2022}{18}}. Therefore,
there is a need for studies that explore spatial differences in
vaccination within the Canadian territory and that consequently, can
help identify disparities that need to be addressed within specific
areas in each province.

This need is especially important in the case of Ontario, the most
populated province of Canada. Between 2007 and 2019, Ontario managed
healthcare access to its inhabitants using 14 intra-provincial divisions
called the Local Health Integration Networks (LHINs), which aimed to
provide an integrated health system for the province. However, this
approach was complex, bureaucratic, and resulted in excessive
expenditures, disparities in mortality rates, the deterioration of
certain performance indicators such as wait times and hospital
readmissions, fragmented electronic health systems, the decline of
performance indicators, and inequities in health services
access\textsuperscript{\protect\hyperlink{ref-tsasis2012}{33}--\protect\hyperlink{ref-lysyk2016}{37}}.
Therefore, with the intent of better organizing and delivering care in
late 2019 the provincial government eliminated the LHINs and
incorporated the areas covered by them into six larger Health Regions
(North East, North West, Central, Toronto, West, and
East)\textsuperscript{\protect\hyperlink{ref-dong2022}{35}}.

Because of the relatively recent adoption of the Health Region model and
its alignment with the onset of the COVID-19 pandemic, there is a need
to analyze if there are ongoing disparities in health access under this
approach that need to be addressed before they are exploited by a new
disease or public health threat. In this regard, previous research has
highlighted disparities in the level of activity of each Health
Region\textsuperscript{\protect\hyperlink{ref-sethuram2023}{38}}.
Therefore, analyzing differences in vaccination uptake within the Health
Regions and can help identify which socio-demographic groups are the
most vulnerable and what areas of the province deserve special attention
by decision-makers.

Therefore, in this study we aim to understand the differences in
vaccination uptake between the different Health Regions of Ontario
between October of 2021 and January of 2022. By including socio-economic
factors in our analysis, we aimed to identify in which groups these
differences were significant in order to provide an assessment of the
current state of healthcare access in Ontario.

\hypertarget{methods}{%
\section{Methods}\label{methods}}

\hypertarget{sec-data}{%
\subsection{Data and Methods}\label{sec-data}}

We used data from the \emph{Survey of COVID-19 related Behaviours and
Attitudes}, a repeated cross sectional survey focused on the Canadian
province of Ontario that was commissioned by the Fields Institute for
Research in Mathematical Sciences and the Mathematical Modelling of
COVID-19 Task Force under ethical guidance from the University of
Toronto (under protocol 00043317), and which ran between September 30th,
2021 and January 17th,2022. The survey collected socio-economic
information from participants (Table~\ref{tbl-descriptive-stats}), their
location (nearest municipality, as shown in Figure~\ref{fig-map}), the
date of access to the survey, and asked information on vaccination
status by using the question ``Have you received the first dose of the
COVID vaccine?'', with possible answers ``yes'' and ``no''. The original
dataset contained 39,029 observations.

By design, the survey allowed respondents to exit at any time and
deployed the questions randomly. This resulted in \(\approx\) 84\% of
the observations having multiple missing answers or being incomplete due
to participants leaving the survey at different stages. Therefore, we
selected 6,343 observations that were labelled as ``complete'' in the
dataset and that had answers for all covariates considered in our
analysis. Later, we matched the city of each observation with its
corresponding LHIN and Health Region, and removed observations from
areas with low representation (254 observations corresponding to the
North West and North East Health Regions). After all the preliminary
analyses, the total number of observations used for analysis was 6,236
and included the East, Central, Toronto, and West Health Regions
covering between October 1st,2021 and January 17, 2022. The original
dataset, clean dataset, and details on the data cleaning process are
described in detail in the
\href{https://github.com/aimundo/Fields_COVID-19/}{GitHub repository}
for this paper.

\begin{figure}

{\centering \includegraphics[width=1\textwidth,height=1\textheight]{../data/map_data/map_June_01.pdf}

}

\caption{\label{fig-map}Geographic representation of the data collected
by the \emph{Survey of COVID-19 related Behaviours and Attitudes},
collected by the Fields Institute in Ontario. Municipalities from where
survey participants provided answers appear as points, color indicates
number of observation obtained from each city. The Health six Regions
are color-coded and labelled sequentially. Internal boundaries within
certain Health Regions indicate areas previously covered by the Local
Integrated Health Networks (LHINs).}

\end{figure}

\hypertarget{statistical-analyses}{%
\subsection{Statistical analyses}\label{statistical-analyses}}

We used a logistic regression model to examine the impact of the Health
Regions in vaccination rates while considering the socio-economic
factors and months covered by the survey
(Table~\ref{tbl-descriptive-stats}) and certain interactions (Race and
Health Region and Race and income), as previous studies have shown that
socio-economic factors and their interactions are significant predictors
of intent of vaccination and vaccination
status\textsuperscript{\protect\hyperlink{ref-nguyen2022}{39}--\protect\hyperlink{ref-cnat2022a}{41}}.
Because we identified differences in representativity between the survey
data and the estimates from the Census, we used an iterative
proportional fitting procedure
(\emph{raking})\textsuperscript{\protect\hyperlink{ref-deming1940}{42}}
to correct the data using data from the Census and Health Region
population totals; and fitted the regression model to the uncorrected
and corrected data. Details regarding the correction can be found in the
Appendix. All analyses were conducted in R 4.2.2 using the packages
\texttt{survey}\textsuperscript{\protect\hyperlink{ref-lumley2011}{43}},\texttt{tidyverse}\textsuperscript{\protect\hyperlink{ref-wickham2019}{44}},
\texttt{quarto}\textsuperscript{\protect\hyperlink{ref-quarto}{45}},
\texttt{modelsummary}\textsuperscript{\protect\hyperlink{ref-modelsummary}{46}},
and
\texttt{gtsummary}\textsuperscript{\protect\hyperlink{ref-gtsummary}{47}}.

\hypertarget{results}{%
\section{Results}\label{results}}

\hypertarget{sample-characteristics}{%
\subsection{Sample Characteristics}\label{sample-characteristics}}

Table~\ref{tbl-descriptive-stats} shows the characteristics of the data
from the Fields COVID-19 survey used for analysis. The sample contained
6,236 observations, from which 24.8\% (1,547) corresponded to
individuals that reported not having received the first dose of the
vaccine. Vaccination rates ranged between 71-79\% across all household
income brackets, age groups, Health Regions, and the months considered
in the survey. However, the highest vaccination rates in each category
were reported by individuals in the highest income bracket (79\%), those
between 16 and 34 years of age (77\%), individuals that lived in the
East Health Region (77\%), and during January of 2022 (78\%). Between
racial/ethnic groups, the highest vaccination rate was reported by
White/Caucasian individuals (84\%), against vaccination rates between
63-66\% reported in the case of Arab/Middle Eastern, Black, Indigenous,
Latin American individuals, and those that reported belonging to
``Other'' racial groups, which included Southeast Asian, Filipino, West
Asian, and minorities not identified elsewhere.

\hypertarget{tbl-descriptive-stats}{}
\setlength{\LTpost}{0mm}
\begin{longtable}{lccc}
\caption{\label{tbl-descriptive-stats}Descriptive Statistics of the Fields COVID-19 Survey (by Vaccination
Status) }\tabularnewline

\toprule
\textbf{Variable} & \textbf{no}, N = 1,547\textsuperscript{\textit{1}} & \textbf{yes}, N = 4,689\textsuperscript{\textit{1}} & \textbf{p-value}\textsuperscript{\textit{2}} \\ 
\midrule
Income (CAD) &  &  & <0.001 \\ 
    60000 and above & 542 (21\%) & 1,996 (79\%) &  \\ 
    25000-59999 & 347 (25\%) & 1,046 (75\%) &  \\ 
    under 25000 & 658 (29\%) & 1,647 (71\%) &  \\ 
Age Group &  &  & 0.002 \\ 
    16-34 & 645 (23\%) & 2,117 (77\%) &  \\ 
    35-54 & 411 (24\%) & 1,305 (76\%) &  \\ 
    55 and over & 491 (28\%) & 1,267 (72\%) &  \\ 
Health Region &  &  & 0.3 \\ 
    Toronto & 593 (26\%) & 1,709 (74\%) &  \\ 
    Central & 372 (26\%) & 1,083 (74\%) &  \\ 
    East & 236 (23\%) & 783 (77\%) &  \\ 
    West & 346 (24\%) & 1,114 (76\%) &  \\ 
Month &  &  & <0.001 \\ 
    October & 469 (27\%) & 1,263 (73\%) &  \\ 
    November & 376 (28\%) & 980 (72\%) &  \\ 
    December & 181 (24\%) & 565 (76\%) &  \\ 
    January & 521 (22\%) & 1,881 (78\%) &  \\ 
Race &  &  & <0.001 \\ 
    White/Caucasian & 354 (16\%) & 1,871 (84\%) &  \\ 
    Arab/Middle Eastern & 111 (34\%) & 220 (66\%) &  \\ 
    Black & 159 (34\%) & 303 (66\%) &  \\ 
    East Asian/Pacific Islander & 94 (19\%) & 404 (81\%) &  \\ 
    Indigenous & 112 (37\%) & 194 (63\%) &  \\ 
    Latin American & 99 (34\%) & 195 (66\%) &  \\ 
    Mixed & 177 (30\%) & 411 (70\%) &  \\ 
    Other\textsuperscript{\textit{3}} & 315 (34\%) & 606 (66\%) &  \\ 
    South Asian & 126 (21\%) & 485 (79\%) &  \\ 
\bottomrule
\end{longtable}
\begin{minipage}{\linewidth}
\textsuperscript{\textit{1}}n (\%)\\
\textsuperscript{\textit{2}}Pearson's Chi-squared test\\
\textsuperscript{\textit{3}}Southeast Asian, Filipino, West Asian,
and minorities not identified elsewhere according to the Census.\\
\end{minipage}

\hypertarget{multivariate-regression}{%
\subsection{Multivariate Regression}\label{multivariate-regression}}

Figure~\ref{fig-models} presents the estimates (as odd ratios) from the
logistic regression models for vaccination status using the
socio-demographic factors collected by the survey, and their
interactions. Generally speaking, lower odds of vaccination were
identified in both cases in individuals characterized by a low household
income, or that identified as part of underrepresented groups. However,
the magnitude of the estimates differed between the uncorrected and
corrected models and more importantly, there were differences in the
statistical significance of certain estimates before and after the
correction. Specifically, the uncorrected model showed significant
differences in vaccination odds between the age groups considered, the
East Health Region, Latin American individuals with a household income
under CAD 25,000, and Indigenous individuals living in the Central
Health Region (Figure~\ref{fig-models},B) but these were deemed non
statistically significant after the correction.

\begin{figure}

\includegraphics{main_files/figure-pdf/fig-models-1.pdf} \hfill{}

\caption{\label{fig-models}Coefficient estimates and confidence
intervals for the uncorrected model. Only statistically significant
interaction terms are shown. Full interaction terms can be found in
Supplementary Figures A-3 and A-4.}

\end{figure}

However, significantly lower odds of vaccination were identified in the
corrected model for those with a household income under CAD 25,000
(OR=0.37, CI={[}0.27,0.51{]}) and those with an income between CAD
25,000 and 59,999 (OR=0.58, CI={[}0.42,0.81{]}). Additionally,
individuals who identified as Arab/Middle Eastern, Black, Latin
American, of mixed background, or that belonged to other racial groups
(a category that included Southeast Asian, Filipino, West Asian, and
minorities not identified elsewhere), had significantly lower odds of
vaccination than those in the White/Caucasian group (ORs and CIs=0.28
{[}0.16,0.51{]}, 0.27 {[}0.16,0.45{]}, 0.40 {[}0.21,0.76{]}, 0.53
{[}0.30,0.92{]}, 0.23 {[}0.15,0.36{]}). Additionally, individuals that
reported living in the Central and West Health Regions had higher odds
of vaccination than those in the Health Region of Toronto (ORs and
CIs=1.61 {[}1.10,2.34{]}, and 1.59 {[}1.16,2.19{]}, respectively).

Interestingly, individuals in underrepresented groups with a household
income below CAD 25,000 had higher odds of vaccination (when compared to
those with a household income above CAD 60,000). This held true in the
case of Arab/Middle Eastern individuals (OR=34, CI={[}1.70,6.79{]}),
Black individuals (OR=3.81, CI={[}2.05, 7.09{]}), and those in other
racial or ethnic groups (OR=3.19, CI={[}2.00,5.09{]}). Additionally,
individuals with an income between CAD 25,000 and 59,999 in the
Arab/Middle Eastern and other racial or ethnic groups also had higher
odds of vaccination than their high-income peers (ORs and CIs=6.96
{[}2.67,18.16{]}, and 3.5 {[}1.85,6.62{]}).

Finally, the place of habitation affected the odds of vaccination for
certain underrepresented groups, as significantly lower odds of
vaccination were identified for the interaction between Health Region
and race in the case of Black individuals in the Central Health Region
(OR=0.39, CI={[}0.2,0.75{]}), Arab/Middle Eastern individuals in the
East Health Region (OR=0.41 {[}0.17, 0.98{]}), and in the Indigenous and
mixed groups in the West Health Region (ORs and CIs={[}0.31 {[}0.14,
0.7{]} and 0.38 {[}0.19, 0.76{]}, respectively).

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

In this study, we hypothesized that differences in COVID-19 vaccination
uptake were present between the Health Regions between late 2021 and
early 2022. Our goal was to determine which socio-demographic groups
could be impacted by these disparities in order to provide
decision-makers with information that could be used to develop policies
focused on reducing or eliminating these differences and ensuring that
the Health Region model is able to fulfill its mission of improving
health access for all Ontarians.

Our results show that indeed, there were differences in vaccination odds
across Ontario in certain socio-demographic groups. Specifically, those
who identified as Arab/Middle Eastern, Black, Latin American, having
mixed racial or ethnic background, or that belonged to other groups not
explicitly included in the survey (Southeast Asian, Filipino, West
Asian, and minority groups not identified elsewhere) had vaccination
odds that were between a third and a half of that of individuals that
identified as White or Caucasian (Figure~\ref{fig-models}). These
results are consistent with previous studies that have shown lower
vaccination rates in individuals with the same socio-demographic
characteristics\textsuperscript{\protect\hyperlink{ref-guay2022}{18}--\protect\hyperlink{ref-hussain2022}{20},\protect\hyperlink{ref-carter2022}{48}}.

Lower vaccine uptake in the socio-demographic groups indicated above may
be influenced in part, by vaccine hesitancy and refusal, which have been
associated in underrepresented Canadian individuals with concerns on
vaccine safety, effectiveness, and experiences of racial discrimination
in health
settings\textsuperscript{\protect\hyperlink{ref-cnat2022a}{41},\protect\hyperlink{ref-basta2022}{49}--\protect\hyperlink{ref-cnat2023}{51}}.
However, it has been shown that structural barriers also play an
important role in vaccination uptake. In the case of underrepresented
individuals, such barriers include complex scheduling systems, language
barriers, lack of adequate public transportation, and lack of accessible
vaccination
sites\textsuperscript{\protect\hyperlink{ref-njoku2021}{52}}. In this
regard, it is interesting to note that vaccination venues were scarce in
low socio-economic areas that had the highest burden of COVID-19 in
Toronto and other regions of Ontario around the time covered by the
survey\textsuperscript{\protect\hyperlink{ref-bogoch2022}{7},\protect\hyperlink{ref-iveniuk2021}{53}},
and that pharmacies in the Peel region (an area identified as a
``hotspot'' with high numbers of essential workers and multigenerational
households) could not keep up with vaccine
demand\textsuperscript{\protect\hyperlink{ref-gill2022}{54}}. This
suggests that the observed differences are associated with disparities
in vaccine access that were present during the period covered by the
survey.

Interestingly, whereas overall self-reported vaccination rates were
found to be statistically significantly lower in various
underrepresented groups when compared to White/Caucasian individuals,
the change in odds of vaccination within certain racial groups and
income strata was actually positive, in contrast to the White/Caucasian
group, where vaccination odds decreased in income brackets below CAD
60,000 (Supplementary Figure A-5). Specifically, individuals in low
income brackets that belonged to Arab/Middle Eastern, Black, or other
minority groups had higher odds of vaccination that their peers with an
income above 60,000 CAD.

This result is likely reflects in part the fact that individuals in
underrepresented groups tend to perform occupations that have been
deemed as ``essential'' in the context of the
pandemic\textsuperscript{\protect\hyperlink{ref-hawkins2020}{55},\protect\hyperlink{ref-ct2021}{56}},
which include workers in the areas of grocery stores, gas stations,
warehouses, distribution, and manufacturing, all being occupations for
which an income within the significant brackets identified in the
analysis is to be expected. In Ontario, these workers had priority for
COVID-19
vaccination\textsuperscript{\protect\hyperlink{ref-mishra2021}{57}}; and
there is evidence of interventions by vaccination staff in certain parts
of the province to encourage vaccination uptake by these
individuals\textsuperscript{\protect\hyperlink{ref-gill2022}{54}}. These
facts, combined with evidence of increased trends in vaccination in this
group
elsewhere\textsuperscript{\protect\hyperlink{ref-nguyen2021b}{58}},
suggest that the type of occupation of individuals in underrepresented
groups played an important role in increasing the odds of vaccination.

However, the results also indicate that the place of habitation affected
the odds of vaccination for certain underrepresented groups (interaction
term of Health Region and Race, Figure~\ref{fig-models},B).
Specifically, this held true in the case of individuals identifying as
Indigenous or with mixed racial background in the West Health Region,
Black individuals in the Central Health Region, and Arab/Middle Eastern
individuals in the East Health Region Figure~\ref{fig-models}. For these
individuals, vaccination odds were lower when compared to the Toronto
Health Region (Supplementary Figure A-6). We indicate next some
contributing factors that might help provide context to these results.

First, in this case it is useful to analyze the data considering the
LHINs in each Health Region, because most studies in the literature
focused on Ontario use the LHINs as the base of their analyses. The West
Health Region covers the area previously occupied by the Hamilton
Niagara Haldimand Brant, South West, and Waterloo Wellington LHINs,
whereas the East Health Region covers the area of the former Champlain
and Central East LHINs. Previous research has identified health
disparities in these (mostly rural) regions, such as unequal
distribution of primary care providers, increased mortality, and low
pharmacist
availability\textsuperscript{\protect\hyperlink{ref-shah2019}{59}--\protect\hyperlink{ref-timony2022}{61}}.

Furthermore, there is an ongoing challenge for the health system of the
province with regard to personalized healthcare for marginalized
individuals. For example, the West Health Region has only two Aboriginal
Health Access Centres (community-led primary healthcare organizations
focused on First Nations, Métis, and Inuit communities) to provide care
to an estimated 100,000 Indigenous individuals living in the
area\textsuperscript{\protect\hyperlink{ref-ontariohealth}{62}}. Lack of
access to personalized healthcare affects individuals that may mistrust
the traditional healthcare system due to systemic racism or oppression,
which is known to be the case for Indigenous and Black individuals in
Canada, as these rationales have been associated with observed lower
vaccination rates among these
groups\textsuperscript{\protect\hyperlink{ref-smylie2022}{63},\protect\hyperlink{ref-eissa2021}{64}}.
Taken together, this suggests that healthcare disparities specific to
these underrepresented groups in certain parts of the province impacted
vaccination uptake, and highlights the need of investments in the Health
Regions focused on resources, infrastructure, and specially personnel
that can deliver personalized care to marginalized communities, as it
has been shown that such efforts have improved trust in vaccination in
underrepresented groups
elsewhere\textsuperscript{\protect\hyperlink{ref-schafferderoo2020}{65}}.

There are some limitations to the present study. First, the data
collection design, which allowed respondents to withdraw from the survey
at any point, and that deployed the questions in a random manner
resulted in an elevated number of missing observations without a
definite pattern and complicated the implementation of sensitivity
analyses. Therefore, we focused on entries that had complete answers,
and corrected the data using population-wide information from the
Census. More granular corrections would be needed to obtain more
accurate estimates. For example, our analysis identified higher odds of
vaccination in the Central and West Health Regions, but in this case
these differences are likely to be driven by the proportion of
White/Caucasian individuals, who had higher vaccination rates than other
racial groups. Correcting for each racial/ethnic group in each Health
Region can provide a more accurate estimation of region-wide vaccination
rates but unfortunately, at the moment this correction cannot be
implemented as such stratification is has not been implemented in the
Census.

Additionally, our analysis did not consider the North West and North
East Health Regions, due to the low number of entries from these areas
in the survey (Figure~\ref{fig-map}). Low representation is expected as
these regions as they only account for 5\% of the total population of
Ontario. However, these areas have the highest proportion of Indigenous
inhabitants\textsuperscript{\protect\hyperlink{ref-ontariohealth}{62}}.
In the context of personalized care, there is a need to collect data
that focuses on these Health Regions where additional health disparities
might be present and possibly understudied.

The results in this study are based on self-reported data, where bias
might be present. However, in the context of COVID-19, it has been shown
that good agreement exists between self-reported and documented
vaccination
status\textsuperscript{\protect\hyperlink{ref-stephenson2022}{66}}, we
believe that our data was able to provide a valid sample of vaccination
uptake in the province. This is supported by the statistically
significant higher vaccination odds that were identified for January of
2022 in the model, which are consistent with province-wide trends
reported by Public Health Ontario (which show a 4\% increase between
early December and January, in contrast to a 2.5\% increase between
October and
November\textsuperscript{\protect\hyperlink{ref-ontario-covid}{67}});
however, the short time window constitutes essentially a ``snapshot''
view of the evolution of the disease, and additional data would be
needed to obtain estimates per racial/ethnic group over time across all
Health Regions that can help inform the existence of other health
disparities.

Nonetheless, the results presented here can serve as a starting point to
motivate the collection of robust longitudinal data that can be used to
quantify geographical and temporal differences within vulnerable
segments of the population, and that can be used to inform the
development of adequate public health policies within the province of
Ontario or across other provinces in Canada that aim to minimize
disparities in health access.

\hypertarget{conclusion}{%
\section{Conclusion}\label{conclusion}}

@@

\hypertarget{acknowledgments}{%
\section{Acknowledgments}\label{acknowledgments}}

>>>>

# Conclusion

@@
